Global code motion of congruent instructions

### 2017-04-17 paper.tex
fix section 3.4

\GCC{} recently got code-hoisting \cite{GCCCodeHoisting} which is implemented as
part of \GVN{}-\PRE{}: it uses the set of ANTIC\_IN and AVAIL\_OUT value
expressions computed for \PRE{}. ANTIC\_IN[B] contains very busy expressions at
basic block B i.e., values computed on all paths from B to exit and
AVAIL\_OUT[B] contains values which are already available. The algorithm hoists
top down to a predecessor.  It uses ANTIC\_IN[B] to know what expressions will
be computed on every path from B to exit, and can be computed in B.  It uses
AVAIL\_OUT[B] to subtract out those values already being computed.  The cost
function is: for each hoist candidate, if all successors of B are dominated by
B, then we know insertion into B will eliminate all the remaining computations.
It then checks to see if at least one successor of B has the value available.
This avoids hoisting it way up the chain to ANTIC\_IN[B].  It also checks to
ensure that B has multiple successors, since hoisting in a straight line is
pointless.  The algorithm continues top down the dominator tree, working in
tandem with \PRE{} until no more hoisting is possible.  One advantage of \GCC{}
implementation is that it works in sync with the \GVN{}-\PRE{} such that when
new hoisting opportunities are created by \GVN{}-\PRE{}, code-hoisting will
hoist them.

@@

In this example (c1, c2) or (d1, d2) are potential sinkable candidates. Since
(c1, c2) depend on i1 and i2 respectively which are also in their original basic
blocks, c1 and c2 are not anticipable in B3. So without knowing the sinkability
of `i1' and `i2' it would be illegal to sink (c1, c2) to B3. On the other hand
(d1, d2) can safely be sunk because their operands are readily available at the
sink point i.e., B3. It should also be noted that, just because the expressions
are identical and operands are available, it still requires a unique
post-dominating PHI to use the exact same values to be legally sinkable.

@@

\subsection{Profitability check (Cost models)}
\label{sec:cost-models}
After the legality checks have passed, we check if a code-motion is profitable.
That takes into account the impact code-motion would have on various parameters
that might affect runtime performance e.g., impact on live-range, gain in the
code size. Since this is mostly a code-size optimization pass, the goal is to
not regress in performance across popular benchmarks at the same time reduce
code size as much as possible. Following are the cost models which are
implemented:

\subsubsection{Reduce register pressure}
\label{hoist:reg-pressure}
Hoisting upwards will decrease the live-range of its use, if it is a last use (a
kill), but increase the live-range of its definition. Conversely, sinking will
decrease the live-range of the defined register but increase the live-range for
killed operands. If the live-range after code-motion is less than before it will
be moved. Essentially, as long as there is one killed operand , code hoisting
will either decrease or preserve the register pressure.  Similarly, code-sinking
will either decrease or preserve the register pressure as long as there is one
operand killed at most.  Following example explains how code motion of identical
computations can reduce the register pressure.  Consider the following example
where the labels prefixed with 'P' represent the position of instruction in a
basic block (names prefixed with 'B').

@@

B2: a1 = b<kill> + c<kill>

After hoisting a0 and a1 are removed and a copy
of a0 as a01 is placed in B0 just before the branch.
\end{verbatim}

@@

Code-motion basically consists of two parts i.e., hoisting and sinking. This
implementation only moves congruent instructions. An immediate guarantee of this
approach is gain in code size (the final executable may be of larger size
because of more inlining). The algorithm prefers hoisting to sinking. If the
dependency of a hoistable candidate is in the same basic block as the candidate,
then the dependency must also be hoistable otherwise hoisting will be illegal or
would require a complicated code generation to make it legal. The current
algorithm discards cases if the dependency is neither hoistable nor
rematerializable.

@@

Barriers are based on the concept of pinned instructions \cite{click1995global}
but extended to adapt to \LLVM{} IR. Since a basic block in \LLVM{} IR is actually an
extended basic block because there might be calls in the middle of a basic block
which might not return. So barriers can be present in the middle of a basic
block. Essentially, any instruction that cannot not guarantee progress is marked
as a barrier. In the absense of context (as in our current implementation), some
instructions which might be safe are still classified a barrier e.g., volatile
loads/stores, calls with missing attributes.

>>>>

\GCC{} recently got code-hoisting \cite{GCCCodeHoisting} which is implemented as
part of \GVN{}-\PRE{}: it uses the set of ANTIC\_IN and AVAIL\_OUT value
expressions computed for \PRE{}. ANTIC\_IN[B] contains very busy expressions at
basic block B, i.e., values computed on all paths from B to exit and
AVAIL\_OUT[B] contains values which are already available. The algorithm hoists
top down to a predecessor.  It uses ANTIC\_IN[B] to know what expressions will
be computed on every path from B to exit, and can be computed in B.  It uses
AVAIL\_OUT[B] to subtract out those values already being computed.  The cost
function is: for each hoist candidate, if all successors of B are dominated by
B, then we know insertion into B will eliminate all the remaining computations.
It then checks to see if at least one successor of B has the value available.
This avoids hoisting it way up the chain to ANTIC\_IN[B].  It also checks to
ensure that B has multiple successors, since hoisting in a straight line is
pointless.  The algorithm continues top down the dominator tree, working in
tandem with \PRE{} until no more hoisting is possible.  One advantage of \GCC{}
implementation is that it works in sync with the \GVN{}-\PRE{} such that when
new hoisting opportunities are created by \GVN{}-\PRE{}, code-hoisting will
hoist them.

@@

In this example (c1, c2) or (d1, d2) are potential sinkable candidates. Since
(c1, c2) depend on i1 and i2 respectively which are also in their original basic
blocks, c1 and c2 are not anticipable in B3. So without knowing the sinkability
of `i1' and `i2' it would be illegal to sink (c1, c2) to B3. On the other hand
(d1, d2) can safely be sunk because their operands are readily available at the
sink point, i.e., B3. It should also be noted that, just because the expressions
are identical and operands are available, it still requires a unique
post-dominating PHI to use the exact same values to be legally sinkable.

@@

\subsection{Profitability check (Cost models)}
\label{sec:cost-models}
After the legality checks have passed, we check if a code-motion is profitable.
That takes into account the impact code-motion would have on various parameters
that might affect runtime performance, e.g., impact on live-range, gain in the
code size. Since this is mostly a code-size optimization pass, the goal is to
not regress in performance across popular benchmarks and at the same time to reduce
code size as much as possible. Following are the cost models which are
implemented:

\subsubsection{Reduce register pressure}
\label{hoist:reg-pressure}
Hoisting upwards will decrease the live-range of its use, if it is a last use (a
kill), but increase the live-range of its definition. Conversely, sinking will
decrease the live-range of the defined register but increase the live-range for
killed operands. If the live-range after code-motion is less than before it will
be moved. Essentially, as long as there is one killed operand, code hoisting
will either decrease or preserve the register pressure.  Similarly, code-sinking
will either decrease or preserve the register pressure as long as there is one
operand killed at most.  Following example explains how code motion of identical
computations can reduce the register pressure.  Consider the following example
where the labels prefixed with 'P' represent the position of instruction in a
basic block (names prefixed with 'B').

@@

B2: a1 = b<kill> + c<kill>
\end{verbatim}

After hoisting a0 and a1 are removed and a copy
of a0 as a01 is placed in B0 just before the branch.

@@

Code-motion basically consists of two parts, i.e., hoisting and sinking. This
implementation only moves congruent instructions. An immediate guarantee of this
approach is gain in code size (the final executable may be of larger size
because of more inlining). The algorithm prefers hoisting to sinking. If the
dependency of a hoistable candidate is in the same basic block as the candidate,
then the dependency must also be hoistable otherwise hoisting will be illegal or
would require a complicated code generation to make it legal. The current
algorithm discards cases if the dependency is neither hoistable nor
rematerializable.

@@

Barriers are based on the concept of pinned instructions \cite{click1995global}
but extended to adapt to \LLVM{} IR. Since a basic block in \LLVM{} IR is actually an
extended basic block because there might be calls in the middle of a basic block
which might not return. So barriers can be present in the middle of a basic
block. Essentially, any instruction that cannot not guarantee progress is marked
as a barrier. In the absense of context (as in our current implementation), some
instructions which might be safe are still classified a barrier, e.g., volatile
loads/stores, calls with missing attributes.
